A content-sensitive framework for modeling episodic memory reveals event-like structure in naturalistic experience, recall, and neural processing

### 2019-08-29 paper/main.tex
updated fig 6 text

Because our analysis framework projects the dynamic video content and participants' recalls onto a shared topic space, and because the dimensions of that space are known (i.e., each topic dimension is a set of weights over words in the vocabulary; Fig.~\topics), we can examine the topic trajectories to understand which specific content was remembered well (or poorly).  For each video event, we can ask: what was the average correlation (across participants) between the video event's topic vector and the closest matching recall event topic vectors from each participant? This yields a single correlation coefficient for each video event, describing how closely participants' recalls of the event tended to reliably capture its content (Fig.~\ref{fig:topics}A).  (We also examined how different comparisons between each video event's topic vector and the corresponding recall event topic vectors related to hand-annotated characterizations of memory performance; see \textit{Supporting Information}).  Given this summary of which events were recalled reliably (or not), we next asked whether the better-remembered or worse-remembered events tended to reflect particular topics.  We computed a weighted average of the topic vectors for each video event, where the weights reflected how reliably each event was recalled.  To visualize the result, we created a ``wordle'' image~\citep{MuelEtal18} where words weighted more heavily by better-remembered topics appear in a larger font (Fig.~\ref{fig:topics}B, green box).  Events that reflected topics weighting heavily on characters like ``Sherlock'' and ``John'' (the main characters) and locations like ``221b Baker Street'' (a major recurring location and the address of the flat that Sherlock and John share) were best remembered.  An analogous analysis revealed which themes were poorly remembered.  Here in computing the weighted average over events' topic vectors, we weighted each event in \textit{inverse} proportion to how well it was remembered (Fig.~\ref{fig:topics}B, red box).  This revealed that events with relatively minor characters such as ``Mike,'' ``Jeffrey,'' and ``Molly,'' as well as less-integral plot locations (e.g., ``hospital'' and ``office'') were least well-remembered.  This suggests that what is retained in memory are the major plot elements (i.e., the overall shape of what happened), whereas the more minor details are prone to pruning.

@@

In addition to constructing overall summaries, assessing the video and recall topic vectors from individual events can provide further insights.  Specifically, for any given event we can construct two wordles: one from the original video event's topic vector, and a second from the average topic vectors produced by all participants' recalls of that event.  We can then examine those wordles visually to gain an intuition for which aspects of the video event were recapitulated in participants' recalls.  Several example wordles are displayed in Figure~\ref{fig:topics}C (wordles from the three best-remembered events are circled in green; wordles from the three worst-remembered events are circled in red).  Using wordles to visually compare the topical content of each video event and the (average) corresponding recall event reveals the specific content from the specific events that is reliably retained in the transformation into memory (green events) or not (red events).

@@

Perhaps the simplest and most widely used measure of memory performance is \textit{accuracy}---i.e., the proportion of studied (experienced) items (in this case, the 34 video events) that the participant later remembered.  \cite{ChenEtal17} developed a human rating system whereby the quality of each participant's memory was evaluated by an independent rater.  We found a strong across-participants correlation between these independant ratings and the overall number of events that our HMM approach identified in participants' recalls (Pearson's $r(15) = 0.67, p = 0.003$).

@@

\paragraph*{Precision.}
We tested whether participants who recalled more events were also more \textit{precise} in their recollections. For each participant, we computed the correlation between the topic vectors for each recall event and that of its closest-matching video event (only for the events which they recalled). We Fisher's $z$-transformed the correlations, computed the average and then inverse Fisher's z-transformed the resulting value.  This gave a single value per participant representing the average precision across all recalled events.  We then correlated this value with hand-annotated as well as model derived (e.g. $k$ or the number of events recovered by the HMM) memory performance.

\paragraph*{Distinctiveness.}
We also considered the \textit{distinctiveness} of each recalled event. That is, how uniquely a recalled event's topic vector matched a given video event topic vector, versus the topic vectors for the other video events. We hypothesized that participants with high memory performance might describe each event in a more distinctive way (relative to those with lower memory performance who might describe events in a more general way).  To test this hypothesis we define a distinctiveness score for each recalled event as

%NOTE: Need to check on this equation!
\[
  d(\mathrm{event}) = 1 - \bar{c}(\mathrm{event}),
\]
where $\bar{c}(\mathrm{event})$ is the average correlation between the given recalled event's topic vector and the topic vectors from all video events \textit{except} the best-matching video event.  We then averaged these distinctiveness scores across all of the events recalled by the given participant. As above, we used Fisher's $z$-transform (and inverse-transform) before/after averaging correlation values. Finally, we correlated these values with hand-annotated and model derived memory performance scores across-subjects.

@@

\subsubsection*{Estimating the consistency of flow through topic space across participants}
In Figure~\ref{fig:trajectory}B, we present an analysis aimed at characterizing locations in topic space that different participants move through in a consistent way (via their recall topic trajectories).  The two-dimensional topic space used in our visualizations (Fig.~\ref{fig:trajectory}) ranged from -5 to 5 (arbitrary) units in the $x$ dimension and from -6.5 to 2 units in the $y$ dimension.  We divided this space into a grid of vertices spaced 0.25~units apart.  For each vertex, we examined the set of line segments formed by connecting each pair successively recalled events, across all participants, that passed within 0.5 units.  We computed the distribution of angles formed by those segments and the $x$-axis, and used a Rayleigh test to determine whether the distribution of angles was reliably ``peaked'' (i.e., consistent across all transitions that passed through that local portion of topic space).  To create Figure~\ref{fig:trajectory}B we drew an arrow originating from each grid vertex, pointing in the direction of the average angle formed by line segments that passed within 0.5 units.  We set the arrow lengths to be inversely proportional to the $p$-values of the Rayleigh tests at each vertex.  Specifically, for each vertex we converted all of the angles of segments that passed within 0.5 units to unit vectors, and we set the arrow lengths at each vertex proportional to the length of the (circular) mean vector.  We also indicated any significant results ($p < 0.05$, corrected using the Benjamani-Hochberg procedure) by coloring the arrows in blue (darker blue denotes a lower $p$-value, i.e., a longer mean vector); all tests with $p \geq 0.05$ are displayed in gray and given a lower opacity value.

>>>>

Because our analysis framework projects the dynamic video content and participants' recalls onto a shared topic space, and because the dimensions of that space are known (i.e., each topic dimension is a set of weights over words in the vocabulary; Fig.~\topics), we can examine the topic trajectories to understand which specific content was remembered well (or poorly).  For each video event, we can ask: what was the average correlation (across participants) between the video event's topic vector and the closest matching recall event topic vectors from each participant? This yields a single correlation coefficient for each video event, describing how closely participants' recalls of the event tended to reliably capture its content (Fig.~\ref{fig:topics}A).  (We also examined how different comparisons between each video event's topic vector and the corresponding recall event topic vectors related to hand-annotated characterizations of memory performance; see \textit{Supporting Information}).  Given this summary of which events were recalled reliably (or not), we next asked whether the better-remembered or worse-remembered events tended to reflect particular topics.  We computed a weighted average of the topic vectors for each video event, where the weights reflected how reliably each event was recalled.  To visualize the result, we created a ``wordle'' image~\citep{MuelEtal18} where words weighted more heavily by better-remembered topics appear in a larger font (Fig.~\ref{fig:topics}B, green box).  Across the full video, content that reflected topics necessary to convey the central focus of the video (e.g., the names of the two main characters, ``Sherlock" and ``John", and the address of a major recurring location, ``221b Baker Street") were best remembered.  An analogous analysis revealed which themes were poorly remembered.  Here in computing the weighted average over events' topic vectors, we weighted each event in \textit{inverse} proportion to how well it was remembered (Fig.~\ref{fig:topics}B, red box).  The least well-remembered video content reflected information not necessary to conveying the video's ``gist," such as the names of relatively minor characters (e.g., ``Mike," ``Jeffrey," ``Molly," and ``Jimmy") and locations (e.g., ``St. Bartholomew's Hospital").

@@

A similar result emerged from assessing the topic vectors for individual video and recall events (Fig.~\ref{fig:topics}C).  Here, for each of the three best- and worst-remembered video events, we have constructed two wordles: one from the original video event's topic vector (left) and a second from the average recall topic vector for that event (right).  The three best-remembered events (circled in green) correspond to scenes important to the central plot-line (Sherlock and John meeting, Sherlock and John chasing the killer, and the killer calling spying on John in a phone booth).  Meanwhile, the three worst-remembered events (circled in red) reflect various side-stories (John talking to his therapist about the war, a soon-to-be-victim's affair with his assistant, and another soon-to-be-victim leaving a party with his friend) that are not essentially to summarizing the video's narrative.

%Events that reflected topics weighting heavily on characters like ``Sherlock'' and ``John'' (the main characters) and locations like ``221b Baker Street'' (a major recurring location and the address of the flat that Sherlock and John share) were best remembered.  An analogous analysis revealed which themes were poorly remembered.  Here in computing the weighted average over events' topic vectors, we weighted each event in \textit{inverse} proportion to how well it was remembered (Fig.~\ref{fig:topics}B, red box).  This revealed that events with relatively minor characters such as ``Mike,'' ``Jeffrey,'' and ``Molly,'' as well as less-integral plot locations (e.g., ``hospital'' and ``office'') were least well-remembered.  This suggests that what is retained in memory are the major plot elements (i.e., the overall shape of what happened), whereas the more minor details are prone to pruning.

% In addition to constructing overall summaries, assessing the video and recall topic vectors from individual events can provide further insights.  Specifically, for any given event we can construct two wordles: one from the original video event's topic vector, and a second from the average topic vectors produced by all participants' recalls of that event.  We can then examine those wordles visually to gain an intuition for which aspects of the video event were recapitulated in participants' recalls.  Several example wordles are displayed in Figure~\ref{fig:topics}C (wordles from the three best-remembered events are circled in green; wordles from the three worst-remembered events are circled in red).  Using wordles to visually compare the topical content of each video event and the (average) corresponding recall event reveals the specific content from the specific events that is reliably retained in the transformation into memory (green events) or not (red events).

@@

Perhaps the simplest and most widely used measure of memory performance is \textit{accuracy}---i.e., the proportion of studied (experienced) items (in this case, the 34 video events) that the participant later remembered.  \cite{ChenEtal17} developed a human rating system whereby the quality of each participant's memory was evaluated by an independent rater.  We found a strong across-participants correlation between these independant ratings and the overall number of events that our HMM approach identified in participants' recalls (Pearson's $r(15) = 0.64, p = 0.006$).

@@

\paragraph*{Precision.}
We tested whether participants who recalled more events were also more \textit{precise} in their recollections. For each participant, we computed the average correlation between the topic vectors for each recall event and those of its closest-matching video event. This gave a single value per participant representing the average precision across all recalled events.  We then Fisher's $z$-transformed these values and correlated them with both hand-annotated and model-derived (i.e., $k$ or the number of events recovered by the HMM) memory performance.

\paragraph*{Distinctiveness.}
We also considered the \textit{distinctiveness} of each recalled event. That is, how uniquely a recalled event's topic vector matched a given video event topic vector, versus the topic vectors for the other video events. We hypothesized that participants with high memory performance might describe each event in a more distinctive way (relative to those with lower memory performance who might describe events in a more general way).  To test this hypothesis we define a distinctiveness score for each recall event as

\[
  d(\mathrm{event}) = 1 - \bar{c}(\mathrm{event}),
\]
where $\bar{c}(\mathrm{event})$ is the average correlation between the given recalled event's topic vector and the topic vectors from all video events \textit{except} the best-matching video event.  We then averaged these distinctiveness scores across all of the events recalled by the given participant. As above, we used Fisher's $z$-transformation before correlating these values with hand-annotated and model derived memory performance scores across-subjects.

@@

\subsubsection*{Estimating the consistency of flow through topic space across participants}
In Figure~\ref{fig:trajectory}B, we present an analysis aimed at characterizing locations in topic space that different participants move through in a consistent way (via their recall topic trajectories).  The two-dimensional topic space used in our visualizations (Fig.~\ref{fig:trajectory}) comprised a 9x9 (arbitrary units) square.  We divided this space into a grid of vertices spaced 0.25~units apart.  For each vertex, we examined the set of line segments formed by connecting each pair successively recalled events, across all participants, that passed within 0.5 units.  We computed the distribution of angles formed by those segments and the $x$-axis, and used a Rayleigh test to determine whether the distribution of angles was reliably ``peaked'' (i.e., consistent across all transitions that passed through that local portion of topic space).  To create Figure~\ref{fig:trajectory}B we drew an arrow originating from each grid vertex, pointing in the direction of the average angle formed by line segments that passed within 0.5 units.  We set the arrow lengths to be inversely proportional to the $p$-values of the Rayleigh tests at each vertex.  Specifically, for each vertex we converted all of the angles of segments that passed within 0.5 units to unit vectors, and we set the arrow lengths at each vertex proportional to the length of the (circular) mean vector.  We also indicated any significant results ($p < 0.05$, corrected using the Benjamani-Hochberg procedure) by coloring the arrows in blue (darker blue denotes a lower $p$-value, i.e., a longer mean vector); all tests with $p \geq 0.05$ are displayed in gray and given a lower opacity value.
